### 2020-06-24 chapters/my-work.tex
Ran grammarly on method

For this Thesis we need constant time complexity to process events one by one in real-time. Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit. Some of the methods discussed also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period. In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream. While in some scenarios this is a much-needed feature, in our use-case it is not. How clear the alerts and insights retrieved from the methods are, is also an important property. Some outlier detection methods use Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.

@@

In Section \ref{sec:stream-superset}, we defined a data stream as a continuous collection of events. Each event is timestamped and contains multiple fields. In the context of this Thesis, each of the event's fields are named features. Our goal is to detect data pattern shifts for each feature in an streaming fashion. Such shifts will be measured between a static in time or reference period versus the observed reality during the streaming phase. Figure \ref{fig:timelines} shows the separated reference and streaming timelines.

@@

To detect data pattern shifts, we do not need to directly compare the entire window contents of the reference and sliding windows. This approach would require storing all window contents hence making the memory consumption grow linearly with target sliding window and/or reference period size. Instead, we want to use a \textit{(a)} fixed cost in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase. Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.

One aggregation that encodes distributions properly are histograms. Histograms are a representation of the distribution of numerical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}. Histograms require storing a fixed-size list of bins and corresponding counts per bin regardless of the volume of data, satisfying property \textit{(a)}. Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and incrementing or decrementing the associated count, respectively, thus satisfying property \textit{(b)}. Recall Section \ref{sec:back-swag-algs} where we discussed generic sliding window aggregation algorithms and remember the Subtract-On-Evict (SOE) algorithm (Algorithm \ref{pseudo:soe}). Computing an histogram aggregation over a sliding window of size $W$ using SOE implies storing all $W$ elements in memory to know the element to remove after a new one is inserted. This violates our constant in memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}. EMA-like histograms computed over sliding windows do not require the storage of window contents. An EMA-like histogram is \textit{(a)} constant in memory, \textit{(b)} maintained in real-time and \textit{(c)} encodes the distribution of each feature values. EMA-like histograms are further explained in Section \ref{sec:ema-hist}.

@@

To compare each feature's reference and target sliding window histogram aggregation we need a measure of divergence. We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein \cite{EncyclopaediaMathematics}. The JSD takes as input two histograms and outputs a value between 0 and 1. The closer this value is to 0 the more similar the distributions are. On the other hand, a JSD value of 1 represents two totally different distributions. We chose to use the JSD metric due to its simplicity and leave as future work other possible tests.

Given a divergence or distance value for the two aggregations (reference and target) we need a measure of normal statistical fluctuations. Since we consider the reference period as the normal state we can find the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis. This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).

@@

We choose to emulate windows with four times the half-life size (or $4n_{1/2}$), meaning we discard events whose contribution is less than 6\%. In fact, we set the half-life to be one quarter of the size of the window we want to emulate. We want to monitor a period extending up to one month (at an error level of about 6\%) so we set the half-life to be the number of tuples that correspond to one week worth of data. This will give more weight to the most recent week but it will include contributions decaying to 6\% in the fourth oldest week.

@@

For each sample, and for each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}). We also add two extra bins that will cover the region to the left and to the right of our histogram, respectively. In other words, we add a bin that covers the region between $-\infty$ and our first bin and another that covers values from the last bin to $+\infty$. This way, observed values outside of the reference period will be placed in these special bins, either left or right.

@@

Each of the histograms is an EMA-like histogram. An EMA-like histogram is essentially a collection of EMA-counts, as described in Section \ref{sec:ema-hist}. Each sample is processed event by event and a discount factor is applied per event, thus building an approximated histogram aggregation, using the reference bins, for each feature of that sample. We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period. We use an approximated histogram to mimic the target approximated histogram we will incrementally maintain in the streaming environment.

@@

Let's assume that the histogram from Figure \ref{fig:skewed-left-distro} encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}. Setting a user-defined threshold of \textit{$\alpha$=0.6} for both features would yield very different results. For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}. Instead we define our threshold as a percentile of the distribution and not as a hard-coded distance constant. For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.

@@

For each feature, we initialize the approximate histogram with the last sample's data, from the batch phase. Then, for each incoming event and for each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}. This is all we do, regarding event processing.

>>>>

For this Thesis, we need constant time complexity to process events one by one in real-time. Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit. Some of the methods discussed also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period. In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream. While in some scenarios this is a much-needed feature, in our use-case it is not. How clear the alerts and insights retrieved from the methods are, is also an important property. Some outlier detection methods use Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.

@@

In Section \ref{sec:stream-superset}, we defined a data stream as a continuous collection of events. Each event is timestamped and contains multiple fields. In the context of this Thesis, each of the event's fields are named features. Our goal is to detect data pattern shifts for each feature in a streaming fashion. Such shifts will be measured between a static in time or reference period versus the observed reality during the streaming phase. Figure \ref{fig:timelines} shows the separated reference and streaming timelines.

@@

To detect data pattern shifts, we do not need to directly compare the entire window contents of the reference and sliding windows. This approach would require storing all window contents hence making the memory consumption grow linearly with target sliding window and/or reference period size. Instead, we want to use a \textit{(a)} fixed cost in-memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase. Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.

One aggregation that encodes distributions properly are histograms. Histograms are a representation of the distribution of numerical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}. Histograms require storing a fixed-size list of bins and corresponding counts per bin regardless of the volume of data, satisfying property \textit{(a)}. Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and incrementing or decrementing the associated count, respectively, thus satisfying property \textit{(b)}. Recall Section \ref{sec:back-swag-algs} where we discussed generic sliding window aggregation algorithms and remember the Subtract-On-Evict (SOE) algorithm (Algorithm \ref{pseudo:soe}). Computing a histogram aggregation over a sliding window of size $W$ using SOE implies storing all $W$ elements in memory to know the element to remove after a new one is inserted. This violates our constant in-memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}. EMA-like histograms computed over sliding windows do not require the storage of window contents. An EMA-like histogram is \textit{(a)} constant in memory, \textit{(b)} maintained in real-time and \textit{(c)} encodes the distribution of each feature values. EMA-like histograms are further explained in Section \ref{sec:ema-hist}.

@@

To compare each feature's reference and target sliding window histogram aggregation we need a measure of divergence. We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein \cite{EncyclopaediaMathematics}. The JSD takes as input two histograms and outputs a value between 0 and 1. The closer this value is to 0 the more similar the distributions are. On the other hand, a JSD value of 1 represents two different distributions. We chose to use the JSD metric due to its simplicity and leave as future work other possible tests.

Given a divergence or distance value for the two aggregations (reference and target) we need a measure of normal statistical fluctuations. Since we consider the reference period as the normal state we can find the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis. This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand-picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).

@@

We choose to emulate windows with four times the half-life size (or $4n_{1/2}$), meaning we discard events whose contribution is less than 6\%. We set the half-life to be one-quarter of the size of the window we want to emulate. We want to monitor a period extending up to one month (at an error level of about 6\%) so we set the half-life to be the number of tuples that correspond to one week worth of data. This will give more weight to the most recent week but it will include contributions decaying to 6\% in the fourth oldest week.

@@

For each sample, and each feature, we compute the approximated histogram using the bins computed for the reference histogram (Figure \ref{fig:ref-hist-bins}). We also add two extra bins that will cover the region to the left and the right of our histogram, respectively. In other words, we add a bin that covers the region between $-\infty$ and our first bin and another that covers values from the last bin to $+\infty$. This way, observed values outside of the reference period will be placed in these special bins, either left or right.

@@

Each of the histograms is an EMA-like histogram. An EMA-like histogram is essentially a collection of EMA-counts, as described in Section \ref{sec:ema-hist}. Each sample is processed event by event and a discount factor is applied per event to each bin (EMA expiration factor), thus building an approximated histogram aggregation, using the reference bins, for each feature of that sample. This procedure is illustrated in Figure \ref{fig:EMA-hist-build}.
\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.8]{figures/ema-hist.png}
      \caption[]{Reference equal-height histogram}
      \label{fig:EMA-hist-build}
    \end{center}
\end{figure}
We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period. We use an approximated histogram to mimic the target approximated histogram we will incrementally maintain in the streaming environment.

@@

Let's assume that the histogram from Figure \ref{fig:skewed-left-distro} encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}. Setting a user-defined threshold of \textit{$\alpha$=0.6} for both features would yield very different results. For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}. Instead, we define our threshold as a percentile of the distribution and not as a hard-coded distance constant. For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.

@@

For each feature, we initialize the approximate histogram with the last sample's data, from the batch phase. Then, for each incoming event and each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}. This is all we do, regarding event processing.

### 2020-06-24 chapters/validation.tex
Ran grammarly on method

We generated multiple synthetic datasets before moving on to real data. Furthermore, we first performed single feature analysis before moving to a multi-feature scenario. The problem of multiple testing (Section \ref{sec:multi-test}) only arises in the latter case so we only apply the Holm-Bonferroni correction (Section \ref{sec:holmbonferroni}) for that one.

>>>>

In order to test our proposed system, we generated multiple synthetic datasets before moving on to real data. As an initial first step we started by validating our system using a single-feature dataset before moving to multi-feature scenarios in order to facilitate system validation considering the complexities of multi-feature datasets and the problem of multiple test correction.
